---
title: "Modeling VEDU response to trait metrics"
author: "Claire Tortorelli"
date: "September 4, 2020"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}

library(tidyr)
library(dplyr)
library(ggplot2)
library(lme4)
library(nlme)
library(tweedie)
library(statmod)
library(splines)
library(cplm)
library(glmmTMB)
library(DHARMa)
library(stringr)
library(multcomp)
library(sjPlot)
library(car)
library(gridExtra)
library(cowplot)
library(here)
```

# Prep data
## read in data
```{r}
# trait x subplot dataframe
traitsdf <- read_csv(here("data","plot.trait.matrix.2020.log.woVD.scaled.after.weighted.average.csv"))

# vedu trait by subplot dataframe
vdtraits <- read_csv(here( "data", "vedu.trait2020.stan.after.weighted.average.csv"))
#rename vedu column to plot_quad for merging later
colnames(vdtraits)[1] <- "plot_quad"


# nearest functional species to vedu distance values by subplot
gapdf <- read_csv(here("data", "plot.trait.gap.log.centered.scaled.csv"))
#rename to have gap (nearest species gap distance to VEDU) in the name
names(gapdf) <- c("plot_quad", "finerootV.totalrootV.gap", "height.gap", "N.perc.gap", "root.shoot.gap", "RootAvgDiam.mm.gap", "RootLength.cm.gap", "SLA.gap")

#biomass data - collected in 2020
bio <- read_csv(here("data", "2020_2019_biomass_cover_data_all_updated_5_4_21.csv"))
#log tranform
bio$logother20.bio <- log(bio$other20.bio)

```

## Merge dataframes
```{r}

traitsdfbio <- merge(traitsdf, bio)

traitsdf <- merge(traitsdfbio, gapdf)

```

## Clean up dataframes
```{r}
#order factors in order of productivity gradient
traitsdf$vegtype <- factor(traitsdf$vegtype, levels = c("ARRI", "ARAR", "SEEP"))

#remove influential outlier <- only species is 12% cover LONU
traitsdf <- traitsdf[which(traitsdf$plot_quad != 'ARAR01_C.1'),]

#convert plot and plot no to factors
traitsdf$plot <- as.factor(traitsdf$plot)
traitsdf$plotno <- as.factor(traitsdf$plotno)

# remove CC plots (unseeded controls) because these were not seeded with VEDU 
traitsdf <- traitsdf[!grepl("CC", traitsdf$plot_quad),]
```

# Model vedu response to community trait metrics
## vedu response to biomass
```{r}
bio.bio <- lme(log(VD2020_biomass) ~  logother20.bio*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

plot(bio.bio)
```

# Vedu response to trait dissimilarity
## Calculate trait dissimilarity (community weighted trait value - VD trait value)
```{r}
#SLA
traitsdf$SLA.dis <- abs(traitsdf$SLA - as.double(vdtraits$SLA))

#root:shoot
traitsdf$root.shoot.dis <- abs(traitsdf$root.shoot - as.double(vdtraits$root.shoot))

#fine:total root V
traitsdf$finerootV.totalrootV.dis <- abs(traitsdf$finerootV.totalrootV - as.double(vdtraits$finerootV.totalrootV))

#height
traitsdf$height.dis <- abs(traitsdf$height - as.double(vdtraits$height))

#root L
traitsdf$RootLength.cm.dis <- abs(traitsdf$RootLength.cm - as.double(vdtraits$RootLength.cm))

#root D
traitsdf$RootAvgDiam.mm.dis <- abs(traitsdf$RootAvgDiam.mm - as.double(vdtraits$RootAvgDiam.mm))

#N perc
traitsdf$N.perc.dis <- abs(traitsdf$N.perc - as.double(vdtraits$N.perc))

```

## Model vedu response to trait dissimilarity (individually by trait)
```{r}
#model VEDU response to trait dissimilarity by trait

#SLA
dis.sla <- lme(log(VD2020_biomass) ~  SLA.dis*vegtype, random = ~ 1|plotno/plot, data = traitsdf)


#root:shoot
dis.rs <- lme(log(VD2020_biomass) ~  root.shoot.dis*vegtype, random = ~ 1|plotno/plot, data = traitsdf)


#rootL
dis.rl <- lme(log(VD2020_biomass) ~ RootLength.cm.dis*vegtype, random = ~ 1|plotno/plot, data = traitsdf)


#rootD
dis.rd <- lme(log(VD2020_biomass) ~  RootAvgDiam.mm.dis*vegtype, random = ~ 1|plotno/plot, data = traitsdf)


#fine:total
dis.fr <- lme(log(VD2020_biomass) ~  finerootV.totalrootV.dis*vegtype, random = ~ 1|plotno/plot, data = traitsdf)


#height
dis.ht <- lme(log(VD2020_biomass) ~  height.dis*vegtype, random = ~ 1|plotno/plot, data = traitsdf)


#N
dis.n <- lme(log(VD2020_biomass) ~ N.perc.dis*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

```

## Check model fits & effect sizes
```{r}
dislist <- list(dis.fr, dis.ht, dis.n, dis.rl, dis.rs, dis.sla)
#plot residuals
lapply(dislist, plot) #residualss appear  normal
#check for interaction effects
lapply(dislist, Anova, type = 3)
#examine effect sizes
lapply(dislist, summary) 
```



# Model community - ventenata directional difference (hierarchical distance)
## calculate difference 
```{r}
#SLA
traitsdf$SLA.dif <- (traitsdf$SLA - as.double(vdtraits$SLA))

#root:shoot
traitsdf$root.shoot.dif <- (traitsdf$root.shoot - as.double(vdtraits$root.shoot))

#fine:total root V
traitsdf$finerootV.totalrootV.dif <- (traitsdf$finerootV.totalrootV - as.double(vdtraits$finerootV.totalrootV))

#height
traitsdf$height.dif <- (traitsdf$height - as.double(vdtraits$height))

#root L
traitsdf$RootLength.cm.dif <- (traitsdf$RootLength.cm - as.double(vdtraits$RootLength.cm))

#root D
traitsdf$RootAvgDiam.mm.dif <- (traitsdf$RootAvgDiam.mm - as.double(vdtraits$RootAvgDiam.mm))

#N
traitsdf$N.perc.dif <- (traitsdf$N.perc - as.double(vdtraits$N.perc))
```


## model vedu response to absolute trait difference individually
```{r}
#SLA
dif.sla <- lme(log(VD2020_biomass) ~  SLA.dif*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

#root:shoot
dif.rs <- lme(log(VD2020_biomass) ~  root.shoot.dif*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

#root L
dif.rl <- lme(log(VD2020_biomass) ~ RootLength.cm.dif*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

#root D
dif.rd <- lme(log(VD2020_biomass) ~  RootAvgDiam.mm.dif*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

#fine:total root V
dif.fr <- lme(log(VD2020_biomass) ~  finerootV.totalrootV.dif*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

#height
dif.ht <- lme(log(VD2020_biomass) ~  height.dif*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

#N
dif.n <- lme(log(VD2020_biomass) ~  N.perc.dif*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

```
## Check model fits & effect sizes
```{r}
diflist <- list(dif.fr, dif.ht, dif.n, dif.rl, dif.rs, dif.sla)
#plot residuals
lapply(diflist, plot) #residualss appear  normal
#check for interaction effects
lapply(diflist, Anova, type = 3)
#examine effect sizes
lapply(diflist, summary) 
```

# Model VEDU response to difference in nearest functional species trait "gap" size 

```{r}
#model vedu response to trait gap separately

gap.sla <- lme(log(VD2020_biomass) ~  SLA.gap*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

gap.rs <- lme(log(VD2020_biomass) ~  root.shoot.gap*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

gap.rl <- lme(log(VD2020_biomass) ~ RootLength.cm.gap*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

gap.rd <- lme(log(VD2020_biomass) ~  RootAvgDiam.mm.gap*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

gap.fr <- lme(log(VD2020_biomass) ~  finerootV.totalrootV.gap*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

gap.ht <- lme(log(VD2020_biomass) ~  height.gap*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

gap.n <- lme(log(VD2020_biomass) ~  N.perc.gap*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

```

## Check model fits & effect sizes
```{r}
gaplist <- list(gap.fr, gap.ht, gap.n, gap.rl, gap.rs, gap.sla)
#plot residuals
lapply(gaplist, plot) #residualss appear  normal
#check for interaction effects
lapply(gaplist, Anova, type = 3)
#examine effect sizes
lapply(gaplist, summary) 
```


# Calculate partial R2 
```{r}
#get all models in a list
mlist <- list(dif.sla, dif.rs, dif.rl, dif.rd, dif.fr,  dif.ht, dif.n,  dis.sla, dis.rs, dis.rl, dis.rd, dis.fr, dis.ht, dis.n, gap.sla, gap.rs, gap.rl, gap.rd, gap.fr,  gap.ht, gap.n, bio.bio)

#name models in list
modelnames <- c('dif.sla', 'dif.rs', 'dif.rl', 'dif.rd', 'dif.fr',  'dif.ht', 'dif.n',  'dis.sla', 'dis.rs', 'dis.rl', 'dis.rd', 'dis.fr', 'dis.ht', 'dis.n', 'gap.sla', 'gap.rs', 'gap.rl', 'gap.rd', 'gap.fr',  'gap.ht', 'gap.n', 'bio.bio')

names(mlist) <- modelnames
names(mlist) #check names
```

## Calculate R2 and AIC for all models
```{r}
library(MuMIn)
library(r2glmm)
library(qpcR)

#function to calculate r2 using Nakagawa and Schielzeth method
r2 <- function(model){
  r2beta(model, method = 'nsj')[1,6]
}


#apply r2 to each model in the model list to extract r2s
# r2 measures how much variation in the data is explained by the model
r2list <- data.frame(unlist(lapply(X = mlist, FUN = r2)))
colnames(r2list) <- "r2" 

#calculate AIC for each model in list
# AIC measures how well the model will fit new data
AIClist <- data.frame(unlist(lapply(mlist, AIC)))
colnames(AIClist) <- "AIC"

#calculate AICc for each model in list
AICclist <- data.frame(unlist(lapply(mlist, AICc)))
colnames(AICclist) <- "AICc"

#calculate AICc model weights
aiccweights <- data.frame(round(Weights(AICclist),6))
colnames(aiccweights) <- "AICcweights"

#combine r2, AIC, AICc and AICc weights dfs
modcomp <- cbind(r2list,AIClist,AICclist,aiccweights)

#label models
modcomp$mod <- modelnames
modcomp$modcat <- str_sub(modcomp$mod, 1,3)

#get weighted average r2, mean r2, and mean AICc
modcomp2 <- modcomp %>%
   group_by(modcat) %>%
   mutate(weighted_r2 = weighted.mean(r2, AICcweights),
          mean_r2 = mean(r2),
          average_AICc = mean(AICc))

#calculate delta AICc
modcomp2$deltaAICc <- modcomp2$AICc - min(modcomp2$AICc)

#write AIC table for 2020 data
#write.csv(modcomp2, "2020AIC_model_trait_table.csv")

```

## Calculate ANOVA for all models & create new df
```{r}
library(data.table)
library(purrr)

#write a function to get anova from model and convert the row name to the first col
getanova <- function (x) {
  setDT(data.frame(Anova(x, type = 3))[4,], keep.rownames = TRUE)[]
}


#apply function to all models and bind to a new df
anovadf <- map_df(mlist, getanova, .id = "model")

#write.csv(anovadf, "2020AnovaTable.csv")

```

# Plot models
Create a function that pulls predicted values and upper and lower confidence intervals for each model
```{r}
#function takes fitted model and empty dataframe
plotlme <- function(fitlme, newdat){
  newdat.lme = newdat
  mod <- deparse(substitute(fitlme))

  preds <- data.frame(prediction=predict(fitlme, 
            newdata = newdat.lme, level = 0))
  
  des = model.matrix(formula(fitlme)[-2], newdat.lme)
  predvar = diag( des %*% vcov(fitlme) %*% t(des) )
  #calculates confidence intervals
  preds$lower = preds$prediction - 2*sqrt(predvar)
  preds$upper = preds$prediction + 2*sqrt(predvar)

  #change col names to reflect predictor
  # colnames(preds) <- c(paste(mod,"predlme"),paste(mod,"lower"), paste(mod, "upper"))
  return(preds)
  
}


```

loop through models and create a long dataframe that combines predicted values, upper, and lower confidence intervals for each model
```{r}
#create empty df
predicted.df <- data.frame(NULL)

#loop through model list and apply plot lme function
for(i in 1:length(mlist)){
  mod <- modelnames[i]
  tempdf <- plotlme(mlist[[i]], traitsdf)
  tempdf$model <- mod

  #append trait values to predictions
  traitvar <-  as.character(mlist[[i]]$terms[[3]])[2] #get name of predictor from model
  tempdf$traitvar <- traitsdf[,which(names(traitsdf) == traitvar)]
  
  predicted.df <- rbind(predicted.df, tempdf)
  
}



```

Prep dataframe for plotting
```{r}
#create dataframe to plot with corrected confidence intervals for lme

#pull out plot_quad, vegtype, & VD2020 biomass
traitsdf.rep <- traitsdf[,c(1,9,13)]
#repeat df to match length of predicted values df
traitsdf.rep <- as.data.frame(apply(traitsdf.rep, MARGIN = 2, rep, length(mlist)))

#bind repeated traits df to predicted df
traitsdf.to.plot.lme <- cbind(traitsdf.rep, predicted.df)

#extract model category (hypothesis) from first 3 characters of the model name
traitsdf.to.plot.lme$modcat <- substr(traitsdf.to.plot.lme$model, 0, 3)
#extract trait name  from last 3 characters of the model name
traitsdf.to.plot.lme$trait <- substr(traitsdf.to.plot.lme$model, 5, 8)

#convert biomass to numeric
traitsdf.to.plot.lme$VD2020_biomass <- as.numeric(as.character(traitsdf.to.plot.lme$VD2020_biomass))

#convert to factor & change factor names from data-collection codes
traitsdf.to.plot.lme$vegtype <- factor(traitsdf.to.plot.lme$vegtype, levels = c("ARRI", "ARAR", "SEEP"), labels = c("scab-flat", "low sage-steppe", "wet meadow"))

#change factor names from data-collection codes
traitsdf.to.plot.lme$trait <- factor(traitsdf.to.plot.lme$trait, levels = c("ht", "n", "sla", "rs", "fr", "rd", "rl", "bio"), labels = c("height", "leaf N",  "SLA", "R:S",  "F:T RootV", "root D", "root L", "biomass" ))

#change factor names from data-collection codes
traitsdf.to.plot.lme$modcat <- factor(traitsdf.to.plot.lme$modcat, levels = c("dis", "gap",  "dif", "bio"), labels = c( "weighted.mean", "nearest.species", "hierarchical.distance", "biomass" ))
```


## Plot hierarchical distance, weighted mean dissimilarity, and nearest species dissimilarity
```{r}
library(ggtext)
#set pallette
pal <- c("#A6761D", "#E6AB02" , "#666666")

# extract traits to plot
traitsonly.to.plot.lme <- traitsdf.to.plot.lme[which(traitsdf.to.plot.lme$modcat == "weighted.mean" | traitsdf.to.plot.lme$modcat == "nearest.species" | traitsdf.to.plot.lme$modcat == "hierarchical.distance"),]


# Make labels for x strips (columns)
col_labels <- c('weighted.mean' = "weighted mean<br>
                dissimilarity:<br>
                r<sup>2</sup>=0.064",
                'nearest.species' = "nearest species<br>
                dissimilarity:<br>
                r<sup>2</sup>=0.036",
                'hierarchical.distance' ="hierarchical<br>
                distance:<br>
                r<sup>2</sup>=0.066")


(p2020traitsonly <- ggplot(traitsonly.to.plot.lme, 
          aes(x = traitvar, y = log(VD2020_biomass),
          color = vegtype, group = vegtype) ) + #select variables
    geom_ribbon( aes(ymin = lower, ymax = upper, fill = vegtype, color = NULL),
          alpha = .15) + #assign ribbon values
    geom_line( aes(y = prediction), size = 1)+ 
    theme_bw( base_size = 12) +
    theme(panel.grid.major = element_blank(), #remove gridlines
          panel.grid.minor = element_blank(),
          strip.background = element_blank(),
          panel.border = element_rect(colour = "black"),
          strip.text.x = element_markdown(),
          legend.title=element_blank())+
    scale_fill_manual(values = pal) + #change values based on palette
    scale_color_manual(values = pal) +
    geom_vline(xintercept = 0, linetype="dotted")+ #add dotted line at 0
    scale_y_continuous(limits = c(-1.5,3.2), breaks = c(-1,0,1,2))+
    guides(color=guide_legend(override.aes=list(fill=NA))) +
    facet_grid(trait ~ modcat, scales = "free", 
          labeller = labeller(.cols = col_labels) ) + #facet on traits and model category
    ylab(expression(italic("V. dubia")~"log biomass")) +
    xlab("standardized trait value"))
#labeller = labeller(modcat = new_labels

#ggsave('p2020traitonly.svg')

```


## Plot VEDU relationship to biomass
```{r}
#select biomass for plotting
biomass.to.plot.lme <- traitsdf.to.plot.lme[which(traitsdf.to.plot.lme$modcat == "biomass"),]
col_labels  <- c('biomass' = "biomass:<br>
                 r<sup>2</sup>=0.077")

#create plot in ggplot
(p2020biomass <- ggplot(biomass.to.plot.lme, aes(x = traitvar, y =
          log(VD2020_biomass),
          color = vegtype, group = vegtype) ) +
    geom_ribbon( aes(ymin = lower, ymax = upper, fill = vegtype, color = NULL),
          alpha = .15) +
    geom_line( aes(y = prediction), size = 1)+
    theme_bw( base_size = 12) +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          strip.background = element_blank(),
          panel.border = element_rect(colour = "black"),
          strip.text.x = element_markdown(),
          legend.title=element_blank())+
    scale_fill_manual(values = pal) +
    scale_color_manual(values = pal) +
    scale_y_continuous(limits = c(-1.5,3.2), breaks = c(-1,0,1,2))+
    guides(color=guide_legend(override.aes=list(fill=NA))) +
    facet_grid(trait ~ modcat, scales = "free", 
          labeller = labeller(.cols = col_labels) ) +
    theme(legend.position = "none")+ #remove legend
    ylab(expression(paste("log (",italic("V. dubia"),"biomass)"))) +
    xlab("log (resident biomass)"))



```


### Combine plots
```{r}
library(ggpubr)

# assign legend
legend <- get_legend(p2020traitsonly +
          theme(legend.text=element_text(size=12)))

#remove legend from traits plot
p2020traitonly2 <- p2020traitsonly + theme(legend.position="none")


(all2020plot <- grid.arrange(p2020traitonly2, 
                      widths=c(4.5, 2), 
                      heights = c(25),
                      arrangeGrob(p2020biomass + ylab(""), 
                      legend, 
                      ncol = 1, nrow = 2, 
                      heights = c(5,10))))

#ggsave("figures/all2020plot.svg", all2020plot)

```