---
title: "Calculating nearest species dissimilarity metrics"
author: "Claire Tortorelli"
date: "September 3, 2020"
output:
  github_document
---

Calculate nearest species dissimilarity for each subplot following Catford et al. 2019


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, eval = FALSE)


library(tidyverse)
library(multcomp)
library(car)
library(data.table)
library(here)
library(robustHD)
```

## Data prep

Read in trait data
```{r data}
#species x traits
traits <- read_csv(here("data", "allTraits_averagedBySpecies_USDAcodes.csv"))

```


standardize the trait matrix "log- transform and standardized

```{r}
#set species to row names to standardize
traits <- column_to_rownames(traits, "species")

# log transform traits
traits.log <- log(traits)

#standardize by trait mean and sd

traits.stan <- standardize(traits.log, centerFun = mean, scaleFun = sd)

```


remove vedu ("ventenata") from traits matrix and create vedu only traits matrix
```{r}
#remove vedu from traits matrix

traits.stan.wovd <- traits.stan[-36,]

#create a vedu only trait matrix
VDtraits <- traits.stan[36,]
#write.csv(VDtraits, "vedu_traits_scaled.and.centered.csv")
```

### Species plot data

read in species plot data
```{r}
#read in plot_species data (plot x species)
plot_sp <- read_csv(here("data","2020_USDAspeciesCover_4traits_seededCommunitySubplots.csv"))
```

Remove rare species (not included in trait matrix) and vedu from community matrix
```{r}
#set plot ID to rownames
plot_sp <- column_to_rownames(plot_sp, "plot_quad")

# Keep only species included in trait matrix - remove rare species that occurred in fewer than 3 subplots
species4anal <- rownames(traits.log)
plot_sp <- plot_sp[ , names(plot_sp) %in% species4anal]

#remove VEDU trait matrix for community only trait matrix
plot_sp.wo.vd <- plot_sp[,-c(36)]
#write.csv(plot_sp.wo.vd, "plot_sp.wo.vd.csv")
```

Convert to long format for calculating difference between VEDU trait values and other species present in the community trait values
```{r}
# add plot_quad is first row
plot_sp.wo.vd$plot_quad <- rownames(plot_sp.wo.vd)

# convert to long format
plot_spwovd.l <- gather(plot_sp.wo.vd, "species", "cover", -plot_quad)

```

## Calculate nearest species distance metric

Calculate difference between each species trait values and vedu trait values
```{r}
vd.dif.traits <- traits.stan.wovd

for (i in 1:dim(vd.dif.traits)[1]){ #1 indicates rows, 2 indicates cols
  vd.dif.traits[i,] = traits.stan.wovd[i,] - as.double(VDtraits)
  }

#assign row names to column 1 and convert to df
vd.dif.traits <- as.data.frame(setDT(as.data.frame(vd.dif.traits), keep.rownames = TRUE)[])
names(vd.dif.traits)[names(vd.dif.traits) == "rn"] <- "species"

#merge differences between vedu and other species trait values by species
diff_plot.merge <- merge(plot_spwovd.l, vd.dif.traits)

#remove species from subplots where cover = 0 (they don't occur in these subplots)
diff_plot.merge0 <- subset(diff_plot.merge, cover > 0)

```


Get trait values just above and below ventenata's
```{r}
#write functions to get the trait values just above and below vedu's
get.min <- function(x) {
min(x[which(x > 0)])}

get.max <- function(x) {
max(x[which(x < 0)])}

# calcualte the trait value just higher than VEDU's
gap_higher <- diff_plot.merge0 %>% group_by(plot_quad) %>%
            summarise_at(c("SLA", "root.shoot",	"RootLength.cm",	"RootAvgDiam.mm",	"finerootV.totalrootV",	"height", "N.perc"), get.min) 

#just lower than VEDU's
gap_lower <- diff_plot.merge0 %>% group_by(plot_quad) %>%
            summarise_at(c("SLA", "root.shoot",	"RootLength.cm",	"RootAvgDiam.mm",	"finerootV.totalrootV",	"height", "N.perc"), get.max) 
                              


gap.h.long <- gather(gap_higher, "trait", "high.value", -plot_quad)
gap.l.long <- gather(gap_lower, "trait", "low.value", -plot_quad)

gap.traits <- cbind(gap.h.long, gap.l.long[,3])
```

### Calculate trait distance to nearest functional species(from VEDU) 
according to Catford et al. (2019) Supplemental material

Calculate absolute value of nearest functional species trait distance to vedu 
```{r}
#convert to long format
VDtraits.long <- gather(VDtraits, trait, vd.value) 
#merge dataframes
gap.traits.wvd <- merge(gap.traits, VDtraits.long)
#calculate absolute value of the trait difference 
gap.traits.wvd$vd.value <- abs(gap.traits.wvd$vd.value) 
```

Calculate nearest functional species trait distance
```{r}
gap.traits.wvd$gap <- NA


for (i in 1:dim(gap.traits.wvd)[1]){ #1 indicates rows, 2 indicates cols
  if (gap.traits.wvd[i,3] == Inf) {
  gap.traits.wvd[i,6] <- abs(gap.traits.wvd[i,4])*2
  } else if ((gap.traits.wvd[i,4] == -Inf) & (gap.traits.wvd[i,3] > gap.traits.wvd[i,5]) ) {
    gap.traits.wvd[i,6] <- abs(gap.traits.wvd[i,3])
  } else if (gap.traits.wvd[i,4] == -Inf & gap.traits.wvd[i,4] <= gap.traits.wvd[i,5] ) {
    gap.traits.wvd[i,6] <- abs(gap.traits.wvd[i,3])*2
  } else {
    gap.traits.wvd[i,6] <- gap.traits.wvd[i,3] - gap.traits.wvd[i,4]
    }
}
  
gap.traits.wide <- spread(gap.traits.wvd[,c(1,2,6)], trait, gap)

```


Write plot_gap dataframe (log, standardized, and scaled prior to nearest species distance ("gap") calculation)
```{r}
#write.csv(gap.traits.wide, "nearest_species_distances.csv", row.names = F)
```



