---
title: "Creating plot x trait weighted average dataframe"
author: "Claire Tortorelli"
date: "September 3, 2020"
output:
  github_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, eval = FALSE)

library(tidyverse)
library(multcomp)
library(car)
library(data.table)
library(here)
```

## Data prep

Read in trait data
```{r data}
#species x traits
traits <- read.csv(here("data","allTraits_averagedBySpecies_USDAcodes.csv"))

```


Log transform trait values to improve normality
```{r}
#set plot ID to rownames
traits <- column_to_rownames(traits, var = "species") 

# log transform trait values
traits.log <- log(traits)

```

Explore correlations in the trait matrix
```{r}

#explore correlations 
cor(as.matrix(traits.log)) 
#rootD and fine:totalRootV are the most strongly correlated: 0.81

#plot correlations
pairs(traits.log, cex.labels=1.5)


```

Remove ventenata from traits matrix & create ventenata only traits matrix
```{r}
#remove vedu from community matrix
traits.log.wo.vd <- traits.log[-36,]

#create a vedu only trait matrix
VDtraits <- traits.log[36,]
#write.csv(VDtraits, "vedu_traits_not.scaled.csv")

```

## Calculate matrix of traits weighted by relative abundance in sublots

Prep matrices for multiplication

Species x Plot matrix prep
```{r}
#read in plot_species data (plot x species)
plot_sp <- read_csv(here("data", "2020_USDAspeciesCover_4traits_seededCommunitySubplots.csv"))

#set plot ID to rownames
plot_sp <- column_to_rownames(plot_sp, var = "plot_quad") 

# Keep only species included in trait matrix - remove rare species that occurred in fewer than 2% of plots
species4anal <- row.names(traits.log)
plot_sp <- plot_sp[ , names(plot_sp) %in% species4anal]

#remove VEDU column from trait matrix for community only trait matrix
plot_sp.wo.vd <- plot_sp[ , !names(plot_sp) %in% c('VEDU')]

#write.csv(plot_sp.wo.vd, "plot_sp.wo.vd.csv")

```

Weight trait matrix by  species relative abundance

multiply plot-species and species- traits matrices
```{r}
# check to see if species are present and in same order
species.in.plot <- colnames(plot_sp)
species.in.plot == species4anal

# multiply matrices
plot.sp.traitmatrix <- data.matrix(plot_sp.wo.vd) %*% data.matrix(traits.log.wo.vd)

```

Divide weighted sum (plot.sp.trait matrix) by total veg cover (without vedu)
```{r}
#create a new dataframe to add total veg cover 
plot_sp.wo.vd.totalveg <- plot_sp.wo.vd

#sum rows to get total veg cover
plot_sp.wo.vd.totalveg$totalveg <- rowSums(plot_sp.wo.vd)

#extract total veg to a new dataframe
totalveg <- data.frame(plot_sp.wo.vd.totalveg[ , names(plot_sp.wo.vd.totalveg) %in% c('totalveg')])

#rename totalveg
colnames(totalveg) <- "totalveg"

#divide trait matrix by total veg
plot.sp.traitmatrix.tveg <- plot.sp.traitmatrix
for (i in 1:dim(plot.sp.traitmatrix)[1]){ #1 indicates rows, 2 indicates cols
  plot.sp.traitmatrix.tveg[i,] = plot.sp.traitmatrix[i,]/as.double(totalveg[i,1])
  }

```

center and scale plot-trait matrix by mean and sd 
(including vedu only "subplot" for comparison to community values)
```{r}

#add ventenata "plot" to matrix with all vd traits
plot.sp.traitmatrix.tveg.vd <- rbind(plot.sp.traitmatrix.tveg, VDtraits)

#center and scale
library(robustHD)
plot.sp.traitmatrix.tveg.vd.stan <- data.frame(standardize(as.matrix(plot.sp.traitmatrix.tveg.vd), centerFun = mean, scaleFun = sd))

#remove vedu row from plot-trait matrix
plot.sp.traitmatrix.tveg.stan.wovd <- plot.sp.traitmatrix.tveg.vd.stan[-151,]

#create a vedu only standardized "plot" value
vedu.trait.stan <- data.frame(plot.sp.traitmatrix.tveg.vd.stan[151,])

#write.csv(vedu.trait.stan, "vedu.trait2020.stan.after.weighted.average.all.cplots.csv")
```


Clean up dataframe for modeling
```{r}
#assign rownames
traitdf <- as.data.frame(setDT(as.data.frame(plot.sp.traitmatrix.tveg.stan.wovd), keep.rownames = TRUE)[])
names(traitdf)[names(traitdf) == "rn"] <- "plot_quad"


#create cols for Vegtype, Plot, Plot_no, and treatment
traitdf$vegtype <- substr(traitdf$plot_quad, 1, 4)
traitdf$plot <- substr(traitdf$plot_quad, 1, 6)
traitdf$plotno <- substr(traitdf$plot_quad, 5, 6)


#write plot sp trait matrix (traits + binary, log, standardized, scaled, and weighted average)
#write.csv(traitdf, "data/Community_weighted_means.csv", row.names = F)
```



