---
title: "Modeling ventenata response to trait metrics"
author: "Claire Tortorelli"
date: "September 4, 2020"
output:
  github_document
---

Examining ventenata biomass response to community metrics

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, eval = FALSE)

library(tidyverse)
library(ggplot2)
library(lme4)
library(nlme)
library(statmod)
library(splines)
library(cplm)
library(multcomp)
library(car)
library(gridExtra)
library(cowplot)
library(here)
```

## Prep data

read in data

Prep CWM dataframe
```{r}
# trait x subplot dataframe
df <- read.csv(here("data","Community_weighted_means.csv"))

#remove ventenata from CWM x subplot dataframe and put into separate dataframe
traitsdf <- df[1:105,]

# ventenata trait by subplot dataframe
vdtraits <- df[106,]
```

Prep nearest species dataframe
```{r}


# nearest functional species to ventenata distance values by subplot
gapdf <- read.csv(here("data", "nearest_species_distances.csv"))

#rename to have gap (nearest species trait distance ("gap") to ventenata's distance) in the name
names(gapdf) <- c("plot_quad", "finerootV.totalrootV.gap", "height.gap", "N.perc.gap", "root.shoot.gap", "RootAvgDiam.mm.gap", "RootLength.cm.gap", "SLA.gap")
```

```{r}
#biomass data - collected in 2020
bio <- read.csv(here("data", "2020_biomass_by_subplot.csv"))
#log transform to improve normality
bio$logother20.bio <- log(bio$resident_g)

```

Merge dataframes
```{r}

traitsdfbio <- merge(traitsdf, bio)

traitsdf <- merge(traitsdfbio, gapdf)

```

Organize dataframe for analysis
```{r}
#order factors in order of productivity gradient
traitsdf$vegtype <- factor(traitsdf$vegtype, levels = c("ARRI", "ARAR", "SEEP"))

#remove outlier that falls outside of the communities type we were aiming to capture (not representative of a low sage-steppe (ARAR) site) <- only species is 12% cover LONU
traitsdf <- traitsdf[which(traitsdf$plot_quad != 'ARAR01_C.1'),]

#convert plot and plot_no to factors
traitsdf$plot <- as.factor(traitsdf$plot)
traitsdf$plotno <- as.factor(traitsdf$plotno)

```

## Model vedu_g response to community trait metrics

### Resident biomass

model ventenata response to resident biomass
```{r}
bio.bio <- lme(log(vedu_g) ~  logother20.bio*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

plot(bio.bio)
```

### Trait dissimilarity

Calculate trait dissimilarity (community weighted trait value - ventenata trait value)
```{r}
#SLA
traitsdf$SLA.dis <- abs(traitsdf$SLA - as.double(vdtraits$SLA))

#root:shoot
traitsdf$root.shoot.dis <- abs(traitsdf$root.shoot - as.double(vdtraits$root.shoot))

#fine:total root V
traitsdf$finerootV.totalrootV.dis <- abs(traitsdf$finerootV.totalrootV - as.double(vdtraits$finerootV.totalrootV))

#height
traitsdf$height.dis <- abs(traitsdf$height - as.double(vdtraits$height))

#root L
traitsdf$RootLength.cm.dis <- abs(traitsdf$RootLength.cm - as.double(vdtraits$RootLength.cm))

#root D
traitsdf$RootAvgDiam.mm.dis <- abs(traitsdf$RootAvgDiam.mm - as.double(vdtraits$RootAvgDiam.mm))

#N perc
traitsdf$N.perc.dis <- abs(traitsdf$N.perc - as.double(vdtraits$N.perc))

```

Model ventenata response to trait dissimilarity (individually by trait)

We chose to model traits individually to maintain interpretability (instead of collapsing axes in a PCA) and maintain reasonable degrees of freedom with our interaction term (rather than combining all traits into one model)
```{r}
#model ventenata response to trait dissimilarity by trait

#SLA
dis.sla <- lme(log(vedu_g) ~  SLA.dis*vegtype, random = ~ 1|plotno/plot, data = traitsdf)


#root:shoot
dis.rs <- lme(log(vedu_g) ~  root.shoot.dis*vegtype, random = ~ 1|plotno/plot, data = traitsdf)


#rootL
dis.rl <- lme(log(vedu_g) ~ RootLength.cm.dis*vegtype, random = ~ 1|plotno/plot, data = traitsdf)


#rootD
dis.rd <- lme(log(vedu_g) ~  RootAvgDiam.mm.dis*vegtype, random = ~ 1|plotno/plot, data = traitsdf)


#fine:total
dis.fr <- lme(log(vedu_g) ~  finerootV.totalrootV.dis*vegtype, random = ~ 1|plotno/plot, data = traitsdf)


#height
dis.ht <- lme(log(vedu_g) ~  height.dis*vegtype, random = ~ 1|plotno/plot, data = traitsdf)


#N
dis.n <- lme(log(vedu_g) ~ N.perc.dis*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

```

Check model fits & effect sizes
```{r}
dislist <- list(dis.fr, dis.ht, dis.n, dis.rl, dis.rs, dis.sla)
#plot residuals
lapply(dislist, plot) #residualss appear  normal
#check for interaction effects
lapply(dislist, Anova, type = 3)
#examine effect sizes
lapply(dislist, summary) 
```



### Hierarchical distance
calculate hierarchical difference 
```{r}
#SLA
traitsdf$SLA.dif <- (traitsdf$SLA - as.double(vdtraits$SLA))

#root:shoot
traitsdf$root.shoot.dif <- (traitsdf$root.shoot - as.double(vdtraits$root.shoot))

#fine:total root V
traitsdf$finerootV.totalrootV.dif <- (traitsdf$finerootV.totalrootV - as.double(vdtraits$finerootV.totalrootV))

#height
traitsdf$height.dif <- (traitsdf$height - as.double(vdtraits$height))

#root L
traitsdf$RootLength.cm.dif <- (traitsdf$RootLength.cm - as.double(vdtraits$RootLength.cm))

#root D
traitsdf$RootAvgDiam.mm.dif <- (traitsdf$RootAvgDiam.mm - as.double(vdtraits$RootAvgDiam.mm))

#N
traitsdf$N.perc.dif <- (traitsdf$N.perc - as.double(vdtraits$N.perc))
```


model ventenata response to absolute trait difference individually 

```{r}
#SLA
dif.sla <- lme(log(vedu_g) ~  SLA.dif*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

#root:shoot
dif.rs <- lme(log(vedu_g) ~  root.shoot.dif*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

#root L
dif.rl <- lme(log(vedu_g) ~ RootLength.cm.dif*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

#root D
dif.rd <- lme(log(vedu_g) ~  RootAvgDiam.mm.dif*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

#fine:total root V
dif.fr <- lme(log(vedu_g) ~  finerootV.totalrootV.dif*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

#height
dif.ht <- lme(log(vedu_g) ~  height.dif*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

#N
dif.n <- lme(log(vedu_g) ~  N.perc.dif*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

```

Check model fits & effect sizes
```{r}
diflist <- list(dif.fr, dif.ht, dif.n, dif.rl, dif.rs, dif.sla)
#plot residuals
lapply(diflist, plot) #residualss appear  normal
#check for interaction effects
lapply(diflist, Anova, type = 3)
#examine effect sizes
lapply(diflist, summary) 
```

### Nearest species dissimilarity 

Model ventenata response to difference in nearest functional species trait "gap" size 

```{r}
#model ventenata response to trait gap separately

gap.sla <- lme(log(vedu_g) ~  SLA.gap*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

gap.rs <- lme(log(vedu_g) ~  root.shoot.gap*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

gap.rl <- lme(log(vedu_g) ~ RootLength.cm.gap*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

gap.rd <- lme(log(vedu_g) ~  RootAvgDiam.mm.gap*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

gap.fr <- lme(log(vedu_g) ~  finerootV.totalrootV.gap*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

gap.ht <- lme(log(vedu_g) ~  height.gap*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

gap.n <- lme(log(vedu_g) ~  N.perc.gap*vegtype, random = ~ 1|plotno/plot, data = traitsdf)

```

Check model fits & effect sizes
```{r}
gaplist <- list(gap.fr, gap.ht, gap.n, gap.rl, gap.rs, gap.sla)
#plot residuals
lapply(gaplist, plot) #residualss appear  normal
#check for interaction effects
lapply(gaplist, Anova, type = 3)
#examine effect sizes
lapply(gaplist, summary) 
```


## Comparing models

Calculate partial R2 for each model
```{r}
#get all models in a list
mlist <- list(dif.sla, dif.rs, dif.rl, dif.rd, dif.fr,  dif.ht, dif.n,  dis.sla, dis.rs, dis.rl, dis.rd, dis.fr, dis.ht, dis.n, gap.sla, gap.rs, gap.rl, gap.rd, gap.fr,  gap.ht, gap.n, bio.bio)

#name models in list
modelnames <- c('dif.sla', 'dif.rs', 'dif.rl', 'dif.rd', 'dif.fr',  'dif.ht', 'dif.n',  'dis.sla', 'dis.rs', 'dis.rl', 'dis.rd', 'dis.fr', 'dis.ht', 'dis.n', 'gap.sla', 'gap.rs', 'gap.rl', 'gap.rd', 'gap.fr',  'gap.ht', 'gap.n', 'bio.bio')

names(mlist) <- modelnames
names(mlist) #check names
```


Extract estimates & CIs into dataframe for each model by each vegetation type

```{r}

library(emmeans)
#biomass
embio <- data.frame(emtrends(bio.bio, ~ logother20.bio | vegtype, 
                       var = "logother20.bio"))
# heirarchical distance
emdif.fr <- data.frame(emtrends(dif.fr, ~ finerootV.totalrootV.dif | vegtype, 
                       var = "finerootV.totalrootV.dif"))
emdif.sla <- data.frame(emtrends(dif.sla, ~ SLA.dif | vegtype, 
                       var = "SLA.dif"))
emdif.ht <- data.frame(emtrends(dif.ht, ~ height.dif | vegtype, 
                       var = "height.dif"))
emdif.rs <- data.frame(emtrends(dif.rs, ~ root.shoot.dif | vegtype, 
                       var = "root.shoot.dif"))
emdif.n <- data.frame(emtrends(dif.n, ~ N.perc.dif | vegtype, 
                       var = "N.perc.dif"))
emdif.rl <- data.frame(emtrends(dif.rl, ~ RootLength.cm.dif | vegtype, 
                       var = "RootLength.cm.dif"))
emdif.rd <- data.frame(emtrends(dif.rd, ~ RootAvgDiam.mm.dif | vegtype, 
                       var = "RootAvgDiam.mm.dif"))

#weighted mean dissimilarity
emdis.fr <- data.frame(emtrends(dis.fr, ~ finerootV.totalrootV.dis | vegtype, 
                       var = "finerootV.totalrootV.dis"))
emdis.sla <- data.frame(emtrends(dis.sla, ~ SLA.dis | vegtype, 
                       var = "SLA.dis"))
emdis.ht <- data.frame(emtrends(dis.ht, ~ height.dis | vegtype, 
                       var = "height.dis"))
emdis.rs <- data.frame(emtrends(dis.rs, ~ root.shoot.dis | vegtype, 
                       var = "root.shoot.dis"))
emdis.n <- data.frame(emtrends(dis.n, ~ N.perc.dis | vegtype, 
                       var = "N.perc.dis"))
emdis.rl <- data.frame(emtrends(dis.rl, ~ RootLength.cm.dis | vegtype, 
                       var = "RootLength.cm.dis"))
emdis.rd <- data.frame(emtrends(dis.rd, ~ RootAvgDiam.mm.dis | vegtype, 
                       var = "RootAvgDiam.mm.dis"))

#nearest species dissimilarity

emgap.fr <- data.frame(emtrends(gap.fr, ~ finerootV.totalrootV.gap | vegtype, 
                       var = "finerootV.totalrootV.gap"))
emgap.sla <- data.frame(emtrends(gap.sla, ~ SLA.gap | vegtype, 
                       var = "SLA.gap"))
emgap.ht <- data.frame(emtrends(gap.ht, ~ height.gap | vegtype, 
                       var = "height.gap"))
emgap.rs <- data.frame(emtrends(gap.rs, ~ root.shoot.gap | vegtype, 
                       var = "root.shoot.gap"))
emgap.n <- data.frame(emtrends(gap.n, ~ N.perc.gap | vegtype, 
                       var = "N.perc.gap"))
emgap.rl <- data.frame(emtrends(gap.rl, ~ RootLength.cm.gap | vegtype, 
                       var = "RootLength.cm.gap"))
emgap.rd <- data.frame(emtrends(gap.rd, ~ RootAvgDiam.mm.gap | vegtype, 
                       var = "RootAvgDiam.mm.gap"))
```

export emtrends results (estimates by vegetation type)
```{r}
#combine all emtrends dfs into list
emtrends_list_all <- list(embio, emdif.fr, emdif.sla, emdif.ht, emdif.rs, emdif.n, emdif.rl, emdif.rd, emdis.fr, emdis.sla, emdis.ht, emdis.rs, emdis.n, emdis.rl, emdis.rd, emgap.fr, emgap.sla, emgap.ht, emgap.rs, emgap.n, emgap.rl, emgap.rd)

#merge dataframe trends dfs after adding column that describes the variable being tested
library(magrittr)
em_trendslist2 <- emtrends_list_all %>%
  map( ~ .x %>%
         mutate(var_name = colnames(.x)[1]) %>%
        set_colnames(c("var_value", "vegtype", "estimate", "SE", "df", "lower.CL", "upper.CL", "var_name")))

#convert list to df
em_trends_df <- do.call(rbind.data.frame, em_trendslist2)  

#exponentiate response
em_trends_df$estimate.exp <- exp(em_trends_df$estimate)

#write csv
#write.csv(em_trends_df[,2:8], "trait_model_estimates_by_vegtype_with_emtrends.csv", row.names =  F)
```


Calculate R2 and AIC for all models
```{r}
library(MuMIn)
library(r2glmm)
library(qpcR)

#function to calculate r2 using Nakagawa and Schielzeth method
r2 <- function(model){
  r2beta(model, method = 'nsj')[1,6]
}


#apply r2 to each model in the model list to extract r2s
# r2 measures how much variation in the data is explained by the model
r2list <- data.frame(unlist(lapply(X = mlist, FUN = r2)))
colnames(r2list) <- "r2" 

#calculate AIC for each model in list
# AIC measures how well the model will fit new data
AIClist <- data.frame(unlist(lapply(mlist, AIC)))
colnames(AIClist) <- "AIC"

#calculate AICc for each model in list
AICclist <- data.frame(unlist(lapply(mlist, AICc)))
colnames(AICclist) <- "AICc"

#calculate AICc model weights
aiccweights <- data.frame(round(Weights(AICclist),6))
colnames(aiccweights) <- "AICcweights"

#combine r2, AIC, AICc and AICc weights dfs
modcomp <- cbind(r2list,AIClist,AICclist,aiccweights)

#label models
modcomp$mod <- modelnames
modcomp$modcat <- str_sub(modcomp$mod, 1,3)

#get weighted average r2, mean r2, and mean AICc
modcomp2 <- modcomp %>%
   group_by(modcat) %>%
   mutate(weighted_r2 = weighted.mean(r2, AICcweights),
          mean_r2 = mean(r2),
          average_AICc = mean(AICc))

#calculate delta AICc
modcomp2$deltaAICc <- modcomp2$AICc - min(modcomp2$AICc)

#write AIC table for 2020 data
#write.csv(modcomp2, "2020AIC_model_trait_table.csv")

```

Perform ANOVA for all models & create new dataframe
```{r}
library(data.table)
library(purrr)

#write a function to get anova from model and convert the row name to the first column
getanova <- function (x) {
  setDT(data.frame(Anova(x, type = 3))[4,], keep.rownames = TRUE)[]
}


#apply function to all models and bind to a new dataframe
anovadf <- map_df(mlist, getanova, .id = "model")

#write.csv(anovadf, "2020AnovaTable.csv")

```

## Plot models

Create a function that pulls predicted values and upper and lower confidence intervals for each model
```{r}
#function takes fitted model and empty dataframe
plotlme <- function(fitlme, newdat){
  newdat.lme = newdat
  mod <- deparse(substitute(fitlme))

  preds <- data.frame(prediction=predict(fitlme, 
            newdata = newdat.lme, level = 0))
  
  des = model.matrix(formula(fitlme)[-2], newdat.lme)
  predvar = diag( des %*% vcov(fitlme) %*% t(des) )
  #calculates confidence intervals
  preds$lower = preds$prediction - 2*sqrt(predvar)
  preds$upper = preds$prediction + 2*sqrt(predvar)

  #change column  names to reflect predictor
  # colnames(preds) <- c(paste(mod,"predlme"),paste(mod,"lower"), paste(mod, "upper"))
  return(preds)
  
}


```

loop through models and create a long dataframe that combines predicted values, upper, and lower confidence intervals for each model
```{r}
#create empty df
predicted.df <- data.frame(NULL)

#loop through model list and apply plot lme function
for(i in 1:length(mlist)){
  mod <- modelnames[i]
  tempdf <- plotlme(mlist[[i]], traitsdf)
  tempdf$model <- mod

  #append trait values to predictions
  traitvar <-  as.character(mlist[[i]]$terms[[3]])[2] #get name of predictor from model
  tempdf$traitvar <- traitsdf[,which(names(traitsdf) == traitvar)]
  
  predicted.df <- rbind(predicted.df, tempdf)
  
}



```

Prep dataframe for plotting
```{r}
#create dataframe to plot with corrected confidence intervals for lme

#pull out plot_quad, vegtype, & VD2020 biomass
traitsdf.rep <- traitsdf[,c(1,11,19)]
#repeat dataframe to match length of predicted values df
traitsdf.rep <- as.data.frame(apply(traitsdf.rep, MARGIN = 2, rep, length(mlist)))

#bind repeated traits dataframe to predicted df
traitsdf.to.plot.lme <- cbind(traitsdf.rep, predicted.df)

#extract model category (hypothesis) from first 3 characters of the model name
traitsdf.to.plot.lme$modcat <- substr(traitsdf.to.plot.lme$model, 0, 3)
#extract trait name  from last 3 characters of the model name
traitsdf.to.plot.lme$trait <- substr(traitsdf.to.plot.lme$model, 5, 8)

#convert biomass to numeric
traitsdf.to.plot.lme$vedu_g <- as.numeric(as.character(traitsdf.to.plot.lme$vedu_g))

#convert to factor & change factor names from data-collection codes
traitsdf.to.plot.lme$vegtype <- factor(traitsdf.to.plot.lme$vegtype, levels = c("ARRI", "ARAR", "SEEP"), labels = c("scab-flat", "low sage-steppe", "wet meadow"))

#change factor names from data-collection codes
traitsdf.to.plot.lme$trait <- factor(traitsdf.to.plot.lme$trait, levels = c("ht", "n", "sla", "rs", "fr", "rd", "rl", "bio"), labels = c("height", "leaf N",  "SLA", "R:S",  "F:T RootV", "root D", "root L", "biomass" ))

#change factor names from data-collection codes
traitsdf.to.plot.lme$modcat <- factor(traitsdf.to.plot.lme$modcat, levels = c("dis", "gap",  "dif", "bio"), labels = c( "weighted.mean", "nearest.species", "hierarchical.distance", "biomass" ))
```


Plot hierarchical distance, weighted mean dissimilarity, and nearest species dissimilarity
```{r}
library(ggtext)
#set pallette
pal <- c("#A6761D", "#E6AB02" , "#666666")

# extract traits to plot
traitsonly.to.plot.lme <- traitsdf.to.plot.lme[which(traitsdf.to.plot.lme$modcat == "weighted.mean" | traitsdf.to.plot.lme$modcat == "nearest.species" | traitsdf.to.plot.lme$modcat == "hierarchical.distance"),]


# Make labels for columns
col_labels <- c('weighted.mean' = "weighted mean<br>
                dissimilarity:",
                'nearest.species' = "nearest species<br>
                dissimilarity:",
                'hierarchical.distance' ="hierarchical<br>
                distance:")


(p2020traitsonly <- ggplot(traitsonly.to.plot.lme, 
          aes(x = traitvar, y = log(vedu_g),
          color = vegtype, group = vegtype) ) + #select variables
    geom_ribbon( aes(ymin = lower, ymax = upper, fill = vegtype, color = NULL),
          alpha = .15) + #assign ribbon values
    geom_line( aes(y = prediction), size = 1)+ 
    theme_bw( base_size = 12) +
    theme(panel.grid.major = element_blank(), #remove gridlines
          panel.grid.minor = element_blank(),
          strip.background = element_blank(),
          panel.border = element_rect(colour = "black"),
          strip.text.x = element_markdown(),
          legend.title=element_blank())+
    scale_fill_manual(values = pal) + #change values based on palette
    scale_color_manual(values = pal) +
    geom_vline(xintercept = 0, linetype="dotted")+ #add dotted line at 0
    scale_y_continuous(limits = c(-1.5,3.2), breaks = c(-1,0,1,2))+
    guides(color=guide_legend(override.aes=list(fill=NA))) +
    facet_grid(trait ~ modcat, scales = "free", 
          labeller = labeller(.cols = col_labels) ) + #facet on traits and model category
    ylab(expression(paste("log (",italic("V. dubia")~" biomass)"))) +
    xlab("standardized trait value"))


#ggsave('p2020traitonly.svg')

```


Plot ventenata relationship to biomass
```{r}
#select biomass for plotting
biomass.to.plot.lme <- traitsdf.to.plot.lme[which(traitsdf.to.plot.lme$modcat == "biomass"),]
col_labels  <- c('biomass' = "biomass:")

#create plot in ggplot
(p2020biomass <- ggplot(biomass.to.plot.lme, aes(x = traitvar, y =
          log(vedu_g),
          color = vegtype, group = vegtype) ) +
    geom_ribbon( aes(ymin = lower, ymax = upper, fill = vegtype, color = NULL),
          alpha = .15) +
    geom_line( aes(y = prediction), size = 1)+
    theme_bw( base_size = 12) +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          strip.background = element_blank(),
          panel.border = element_rect(colour = "black"),
          strip.text.x = element_markdown(),
          legend.title=element_blank())+
    scale_fill_manual(values = pal) +
    scale_color_manual(values = pal) +
    scale_y_continuous(limits = c(-1.5,3.2), breaks = c(-1,0,1,2))+
    guides(color=guide_legend(override.aes=list(fill=NA))) +
    facet_grid(trait ~ modcat, scales = "free", 
          labeller = labeller(.cols = col_labels) ) +
    theme(legend.position = "none")+ #remove legend
    ylab(expression(paste("log (",italic("V. dubia"),"biomass)"))) +
    xlab("log (resident biomass)"))



```


Combine plots
```{r}
library(ggpubr)

# assign legend
legend <- get_legend(p2020traitsonly +
          theme(legend.text=element_text(size=12)))

#remove legend from traits plot
p2020traitonly2 <- p2020traitsonly + theme(legend.position="none")


(all2020plot <- grid.arrange(p2020traitonly2, 
                      widths=c(4.5, 2), 
                      heights = c(25),
                      arrangeGrob(p2020biomass + ylab(""), 
                      legend, 
                      ncol = 1, nrow = 2, 
                      heights = c(5,15))))

#ggsave("figures/all2020plot.svg", all2020plot, height = 8, width = 7.5)

```