---
title: "ventenata response to clearing and vegetation type & test of seeding vs. unseeded control"
author: "Claire Tortorelli"
date: "August 18, 2021"
output:
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)

library(here)
library(tidyverse)
library(ggplot2)
library(lme4)
library(stringr)
library(nlme)
library(emmeans)
```

read in data
```{r data}
bio <- read_csv(here("data", "biomass_data_2019.csv"))
```

### Organize data for analysis
```{r format data}

#add col for plot number, plot, and treatment: C = community, uncleared; N = no neighbors, cleared (apologies for the confusing codes)

bio$plotno <- factor(substr(bio$plot_quad,5,6))
bio$plot <- factor(substr(bio$plot_quad,1,6))
bio$trt <- factor(substr(bio$plot_quad,8,9))

#convert veg type to factor
bio$vegtype <- factor(substr(bio$plot_quad,1,4), levels = c("ARRI", "ARAR" , "SEEP"))


#adjust for zeros
bio$stahel_vd19biomass <- bio$vedu19_g + quantile(bio$vedu19_g, .25)^2/quantile(bio$vedu19_g, .75) 


bio$stahel_vd20biomass <- bio$vedu20_g + 0.01
#decided not to use the Stahel method for 2020 vedu biomass because it added 0.3g which was quite a bit more than the other biomass adjustments (closer to 0.01)

#explore data
bio %>% group_by(vegtype, trt) %>%
  summarise(mean = mean(stahel_vd20biomass))
```

### Fit model

Model VEDU biomass response to removal treatment by veg type
split block design receives random effect for block (plotno) and plot
control for 2019 biomass (indicates potential seed bank and microsite suitability)

```{r}

#model treatment 
mtrt <- lme(log(stahel_vd20biomass) ~ trt + log(stahel_vd19biomass), random = ~ 1|plotno/plot, data = bio)

plot(mtrt) #residuals look symmetrical 
summary(mtrt)

mtrt.emm <- emmeans(mtrt, specs = pairwise ~ trt, type = "response", adjust = "none")

summary(mtrt.emm) 
#no effect of clearing or vegetation type
#large difference between seeded and unseeded subplots


#model treatment  with vegetation interaction
mtrt2 <- lme(log(stahel_vd20biomass) ~ trt*vegtype + log(stahel_vd19biomass), random = ~ 1|plotno/plot, data = bio)

plot(mtrt2) #residuals look symmetrical 
summary(mtrt2)

mtrt.emm2 <- emmeans(mtrt2, specs = pairwise ~ trt|vegtype, type = "response", adjust = "none")

#check for interaction effect
library(car)
Anova(mtrt2, type = 2)

```



### Plot results

Extract means for plotting
```{r}

(means_table = mtrt.emm$emmeans %>%
summary(infer = c(TRUE, FALSE) ) %>%
as.data.frame() )

#rename levels for plotitng




means_table$trt  <- factor(means_table$trt, levels = c("C.", "CC"), labels = c("seeded", "unseeded control"))

```


plot 2020 vedu biomass (post seeding) by treatment (cleared vs. uncleared) and vegetation type
```{r fig.height=4.5, fig.width=4.5}

( g1 = ggplot(means_table,  aes(y = response, x = trt)) +
# Define stock as group this week as well as set x and y axes
geom_point(position = position_dodge(width = .75), size = 3 ) + # Add points, dodge by group
geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL,
linetype = trt, width = 0.5), size = 1,
position = position_dodge(width = .75) ) + # Add errorbars, dodge by group
theme_bw(base_size = 14) +
labs(y = expression(italic("V. dubia")~" mean biomass (g)"))+
scale_color_manual(values = c("#A6761D", "#E6AB02", "#666666"))+
    guides(color = FALSE)+
    scale_y_continuous(breaks = c(2,4,6,8), limits = c(0,8.8))+
theme(legend.direction = "horizontal", # make legend horiz
legend.position = "none", # change legend position
panel.grid.minor = element_blank(),
axis.title.x=element_blank(),# Remove gridlines
panel.grid.major.x = element_blank() ))
```

save plot
```{r}



```



