---
title: "cover and richness by vegtype"
author: "Claire Tortorelli"
date: "10/21/2021"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, eval = FALSE)

library(tidyverse)
library(multcomp)
library(car)
library(data.table)
library(here)
```

read in species plot data
```{r}
#read in plot_species data (plot x species)
plot_sp <- read_csv(here("data", "2020_USDAspeciesCover_4traits_seededCommunitySubplots.csv"))
# biomass data
bio <- read_csv(here("data", "2020_biomass_by_subplot.csv"))
```


### Species richness by vegtype
```{r}

#calculate species richness per subplot

plot_sp$div <- rowSums(plot_sp[,2:40] !=0)
plot_sp$veg <- factor(substr(plot_sp$plot_quad,1,4), levels = c('ARRI', 'ARAR', 'SEEP'))


(species_richness <- plot_sp %>% group_by(veg) %>%
  summarize(mean = mean(div),
             max = max(div),
             min = min(div)))

```


### Cover by vegtype 
```{r}
plot_spnovd <- plot_sp[,-39]

plot_spnovd$tcover <- rowSums(plot_spnovd[,2:39])

plot(plot_spnovd$tcover ~ plot_spnovd$veg)

#average cover
(cover <- plot_spnovd %>% group_by(veg) %>%
  summarize(mean = mean(tcover),
             max = max(tcover),
             min = min(tcover)))
```


model ventenata cover response to vegtype
```{r}
library(glmmTMB)
library(DHARMa)
plot_spnovd$plotno <- as.factor(substr(plot_spnovd$plot_quad,5,6))
plot_spnovd$Plot <- as.factor(substr(plot_spnovd$plot_quad,1,6))

#model with tweedie distribution because cover data has high proportion of zeros
mcov <- glmmTMB(tcover ~ veg + (1|plotno/Plot), family = tweedie, data = plot_spnovd)

plot(simulateResiduals(mcov)) #nice fit

summary(mcov)

```

Compare means with emmeans
```{r}
library(emmeans)
(em <- emmeans(mcov, specs = revpairwise ~ veg, type = "response"))
confint(em)
emcov_df <- data.frame(em$emmeans) 
```

## Model biomass response to vegetation type

prep biomass data
```{r}

bio$plotno <- factor(substr(bio$plot_quad,5,6))
bio$plot <- factor(substr(bio$plot_quad,1,6))

#convert veg type to factor
bio$vegtype <- factor(substr(bio$plot_quad,1,4), levels = c("ARRI", "ARAR" , "SEEP"))

```

```{r}
biosum <- bio %>% group_by(vegtype) %>%
  summarize(mean = mean(resident_g),
            sd = sd(resident_g))

```


```{r}
#random effects for plot and block (plot no)
mbio <- lme(log(resident_g) ~ vegtype, random = ~ 1|plotno/plot, data = bio)
plot(mbio)
summary(mbio)
```

Extract means with emmeans for plotting
```{r}
#use emmeans to reduce 
(embio <- emmeans(mbio, specs = revpairwise ~ vegtype, type = "response", adjust= "none"))
confint(embio)
embio_df <- data.frame(embio$emmeans) 
```

change vegetation from code to full name for plotting
```{r}
bio$vegtype <- factor(bio$vegtype, labels = c("ARRI"="scab-flat", "ARAR"="low sage-steppe", "SEEP" = "wet meadow"))
plot_spnovd$veg <- factor(plot_spnovd$veg, labels = c("ARRI"="scab-flat", "ARAR"="low sage-steppe", "SEEP" = "wet meadow"))
```


### Plot results
plot vedu mean height and confidence intervals at each vegetation tpe
```{r fig.height=3, fig.width=3}
pal <- c("#A6761D", "#E6AB02" , "#666666")

( g1 = ggplot(plot_spnovd, aes(y = tcover, x = veg, color = veg)) +
geom_boxplot()+
theme_bw(base_size = 14) +
labs(y = expression("mean foliar cover (%)"))+
      scale_y_continuous(breaks = c(0,25,50,75,100))+
scale_color_manual(values = pal))+
    guides(color = FALSE)+
theme(panel.grid.minor = element_blank(),
axis.title.x=element_blank(),# Remove gridlines
panel.grid.major.x = element_blank())

plot(plot_spnovd$tcover ~ plot_spnovd$veg)
```

Plot mean resident biomass and 95% CIs at each vegetation type
```{r fig.height=3, fig.width=3}
( g2 = ggplot(bio, aes(y = resident_g, x = vegtype, color = vegtype)) +
geom_boxplot() + 
theme_bw(base_size = 14) +
labs(y = "mean resident biomass (g/subplot)")+
scale_color_manual(values = pal))+
    guides(color = FALSE)+
theme(panel.grid.minor = element_blank(),
axis.title.x=element_blank(),# Remove gridlines
panel.grid.major.x = element_blank())
```

combine figures with cowplot
```{r}
library(cowplot)

svg("foliar_cover_biomass.svg", width = 7, height = 5)

plot_grid(g2 + theme(panel.grid.minor = element_blank(),
axis.title.x=element_blank(),# Remove gridlines
panel.grid.major.x = element_blank()) + guides(color = FALSE), g1 + theme(panel.grid.minor = element_blank(),
axis.title.x=element_blank(),# Remove gridlines
panel.grid.major.x = element_blank()) + guides(color = FALSE), labels = c("(a)", "(b)"))

dev.off()
```