---
title: "species in trait space NMS"
author: "Claire Tortorelli"
date: "September 3, 2019"
output:
  github_document
---


Examining species relationships in trait space using NMS

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, eval = FALSE)


library(tidyverse)
library(vegan)
library(ggplot2)
library(stringr)
library(robustHD)
library(here)
```



## Read in data
```{r data}
traits <- read.csv(here("data","allTraits_averagedBySpecies_USDAcodes.csv"))

#read in abudnance data
abund <- read.csv(here("data","2020_USDAspeciesCover_4traits_seededCommunitySubplots.csv"))


```

## Prep data for NMS

Average cover by species
```{r}
#get m
abund_sum <- abund[,2:40] %>% pivot_longer(cols = 1:39, 
                            names_to = "species", values_to = "abund") %>%
                            group_by(species) %>%
                            summarise(total_cover = sum(abund),
                                      mean_cover = mean(abund),
                                      log_cover = log(abund + 1),
                                     sd_mean = sd(abund))


```


Log transform trait values
```{r}
#Set species to row names
rownames(traits) <- traits$species 
traits$species <- NULL 

# log transform trait values
traits.log <- log(traits)

```

## explore correlations
```{r}
#svg("figures/log_tranformed_raw_traits_corPlot.svg")
#check out correlations
cor(traits.log) 
pairs(traits.log)

#dev.off()
```
Some traits appear not linearly correlated, so using NMDS rather than PCA


scale and center data
```{r}
traits.stan <- standardize(traits.log, centerFun = mean, scaleFun = sd)

```

## Run NMS (species in trait space)

metaMDS can't be run on negative values so shifting to the minimum value so that scores are not negative
```{r}
#shift min so that scores are not negative
minShift <- function(x) {x + abs(min(x))}
traits.stan.shift <- apply(traits.stan, 2, minShift)

#run NMS on shifted data with bray distance measure
traits.stan.shift.NMS <- metaMDS(traits.stan.shift, distance='bray', k=2, trymax=50, autotransform=FALSE, wasscores=TRUE, noshare=FALSE) 
#bray curtis tends to perform better with species and sample similarity
traits.stan.shift.NMS

#plot species ('sites') and traits ('species')
plot(traits.stan.shift.NMS, type='t', display=c('sites', 'species'))

```

```{r}
#extract scores to plot with vectors
variableScores <- data.frame(traits.stan.shift.NMS$species)
sampleScores <- data.frame(traits.stan.shift.NMS$points)

#add species abundances to points dataframe
sampleScores$species <- rownames(sampleScores)
sampleScores_abund <- merge(sampleScores, abund_sum)
```

## Plot NMS
```{r}
#svg("figures/traitnms_scaled_by_abund.svg", height = 5.5, width = 8)

par(family = "sans")
plot(sampleScores_abund$MDS1, sampleScores_abund$MDS2, xlab='NMS 1', ylab='NMS 2', type='n', asp=1, las=1)

points(sampleScores_abund[,2:3], pch=16, cex=(sampleScores_abund$log_cover/2), col="#98DEEC")

points(sampleScores_abund[,2:3], pch=16, col ="#2F76BC", cex = 0.5)


arrows(0, 0, variableScores[, 1], variableScores[, 2], length=0.1, angle=20, col ="#A6761D")
 
#set text
text(variableScores[, 1], variableScores[, 2], pos = 3, offset = 0.5, col = "black", font = 2, rownames(variableScores), cex=0.7)

text(sampleScores_abund$MDS1, sampleScores_abund$MDS2, pos = 1, offset = 0.3, col =  "grey3", font = 1, sampleScores_abund$species,  cex=0.7)

#dev.off()

```

