---
title: "9_3_19_SGH_preliminary_cover"
author: "Claire Tortorelli"
date: "September 3, 2019"
output:
  word_document: default
  html_document: default
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, eval = FALSE)

library(tidyverse)
library(multcomp)
library(car)
library(data.table)
library(here)
library(sjPlot)
library(MuMIn)
library(vegan)
library(nlme)
library(emmeans)
```

read in CWM trait data
```{r data}
cwm <- read.csv(here("data","Community_weighted_means.csv"))
bio <- read_csv(here("data", "2020_biomass_by_subplot.csv"))
```

# Calculate dissimilarity matrix



Compute different distance matrixes for all quads with ventenata only "subplot"

Correlation coefficient from McCune & Grace pg. 47 (ranges from 0 - 1: 0 = completely similar; 1 = completely dissimilar)
```{r}

#convert cmw traits to matrix
cwmSub <- cwm[,1:8]
row.names(cwmSub) <- cwmSub$plot_quad
cwmSub$plot_quad <- NULL

#The cosine of the angle between two vectors is equal to the dot product of this vectors divided by the product of vector magnitude.
#this is standardized by (1-r)/2
get.correlation.coef.mccune <- function(a,b) {
  (1 - (a %*% b/(sqrt(sum(a^2))*sqrt(sum(b^2)))))/2
}


cor.plot.trait.dist <- as.data.frame(as.matrix(cwmSub)[,1])

for (i in 1:nrow(cwmSub)){
  cor.plot.trait.dist[i,] =
    get.correlation.coef.mccune(as.matrix(cwmSub)[i,],as.matrix(cwmSub)[106,])
}

```

Euclidean distance
```{r}

#euclidean: Usual square distance between the two vectors (2 norm).
euc.plot.trait.dist <- as.matrix(vegdist(cwmSub, method = "euclidean"))[-106,106]

#merge distance matrices and remove vedu row with dissimilarity value = 0
dist.matrix.vd.plot <- as.data.frame(cbind(as.matrix(cor.plot.trait.dist)[-106,],euc.plot.trait.dist))


#*low values are most similar 
```


# Model ventenata response to community dissimilarity 

prep df for analysis
```{r}
library(data.table)
#assign row names to column 1 and convert to df
dist.df.vd.plot <- as.data.frame(setDT(as.data.frame(dist.matrix.vd.plot), keep.rownames = TRUE)[])
colnames(dist.df.vd.plot) <- c("plot_quad", "cor.plot.trait.dist", "euc.plot.trait.dist") #and rename columns

#merge vd and trait dfs
vd.trait.df <- merge(dist.df.vd.plot, bio)

#create cols for Vegtype, Plot, and Plot_no
vd.trait.df$vegtype <- substr(vd.trait.df$plot_quad, 1, 4)
vd.trait.df$plot <- substr(vd.trait.df$plot_quad, 1, 6)
vd.trait.df$plotno <- substr(vd.trait.df$plot_quad, 5, 6)

#reorder vegtype factor
vd.trait.df$vegtype <- factor(vd.trait.df$vegtype, levels = c("ARRI", "ARAR" , "SEEP"))

```

Corrleation model
```{r}
#--------------correlation

corcov <- lme(log(vedu_g) ~ cor.plot.trait.dist*vegtype, random = ~ 1|plotno/plot, data = vd.trait.df)

#check residuals
plot(corcov)
summary(corcov)

plot_model(corcov)
plot_model(corcov, type = "pred")
```


Euclidean model
```{r}
#--------euclidean distance
euccov <- lme(log(vedu_g) ~ euc.plot.trait.dist*vegtype, random = ~ 1|plotno/plot, data = vd.trait.df)

#check residuals
plot(euccov) #look good

summary(euccov)

#AICc
AICc(euccov)

#plot the response
plot_model(euccov)
plot_model(euccov, type = "pred")


emveg <- data.frame(emtrends(euccov, ~ euc.plot.trait.dist | vegtype, 
                       var = "euc.plot.trait.dist"))

#calcualte R2
library(MuMIn)
library(r2glmm)

#calculate r2 using Nakagawa and Schielzeth method
r2beta(euccov, method = 'nsj')[1,6]

```




### Plot

Plot model and 95% CI from https://aosmith.rbind.io/2018/11/16/plot-fitted-lines/#confidence-intervals-for-lme-objects
```{r}
#create new df
newdat.lme = data.frame(vegtype = vd.trait.df$vegtype,
                        vedu_g = vd.trait.df$vedu_g,
                        euc.plot.trait.dist = vd.trait.df$euc.plot.trait.dist)

#add predicted
newdat.lme$predlme <- predict(euccov, newdata = newdat.lme, level = 0)


#get confidence intervals for predictions
des = model.matrix(formula(euccov)[-2], newdat.lme)
predvar = diag( des %*% vcov(euccov) %*% t(des) )
newdat.lme$lower = with(newdat.lme, predlme - 2*sqrt(predvar) )
newdat.lme$upper = with(newdat.lme, predlme + 2*sqrt(predvar) )

#rename vegtype
newdat.lme$vegtype <- factor(newdat.lme$vegtype, labels = c("ARRI"="scab-flat", "ARAR"="low sage-steppe", "SEEP" = "wet meadow"))
```

```{r}
pal <- c("#A6761D", "#E6AB02" , "#666666")

(pEuc <- ggplot(newdat.lme, aes(x = euc.plot.trait.dist, y = log(vedu_g), color = vegtype) ) +
    geom_ribbon( aes(ymin = lower, ymax = upper, fill = vegtype, color = NULL),
          alpha = .15) + #assign ribbon values
      geom_line(aes(y = predlme), size = 1.5)+ 
    theme_bw( base_size = 12) +
    theme(panel.grid.major = element_blank(), #remove gridlines
          panel.grid.minor = element_blank(),
          strip.background = element_blank(),
          panel.border = element_rect(colour = "black"),
          legend.title=element_blank(),
          legend.position = c(.23,.82))+
  scale_color_manual(values = pal)+
  scale_fill_manual(values = pal)+
  ylab(expression(paste("log (",italic("V. dubia")," biomass)"))) +
    xlab("Euclidean distance"))
```
Save plot
```{r}
library(svglite)
#ggsave("eucDist.svg", pEuc, width = 4, height = 4)
```

# plot Euclidean distances
```{r}
#rename vegtype
vd.trait.df$vegtype <- factor(vd.trait.df$vegtype, labels = c("ARRI"="scab-flat", "ARAR"="low sage-steppe", "SEEP" = "wet meadow"))

ggplot(data = vd.trait.df, aes(x = vegtype, y = euc.plot.trait.dist, color = vegtype)) +
         geom_boxplot() +
         geom_point(size = 2, alpha = .5) +
             theme_bw( base_size = 12) +
    theme(panel.grid.major = element_blank(), #remove gridlines
          panel.grid.minor = element_blank(),
          strip.background = element_blank(),
          panel.border = element_rect(colour = "black"),
          legend.title=element_blank(),
          legend.position = "none")+
  scale_color_manual(values = pal)+
  ylab("Euclidean distance") +
    xlab("")


```





